Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-03-10 mpaper/paper.tex
Dependency graph for the theoretical results

\documentclass{mpaper}
\usepackage{bm}
\usepackage{bbm}
\usepackage{mathtools}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}

@@

@@

% TODO: add info from slides/handout
\section{Introduction}
% problem description/statement

% TODO: will need to rewrite this to avoid plagiarism
Inverse reinforcement learning (IRL)---a problem proposed by Russell in
1998 \cite{DBLP:conf/colt/Russell98}---asks us to find a reward function for a
Markov decision process (MDP) that best explains a set of given demonstrations.
IRL is important because reward functions can be hard to define manually
\cite{DBLP:conf/icml/PieterN04,DBLP:journals/corr/abs-1806-06877}, and rewards
are not entirely specific to a given environment, allowing one to reuse the same
reward structure in previously unseen environments
\cite{DBLP:journals/corr/abs-1806-06877,DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}.
Moreover, IRL has seen a wide array of applications in autonomous vehicle
control \cite{DBLP:journals/ijsr/KimP16,DBLP:journals/ijrr/KretzschmarSSB16} and
learning to predict another agent's behaviour
\cite{DBLP:journals/ai/BogertD18,DBLP:conf/aaai/VogelRGR12,ziebart2008maximum,DBLP:conf/huc/ZiebartMDB08,DBLP:conf/iros/ZiebartRGMPBHDS09}.
Most approaches in the literature (see Section \ref{sec:background}) make a
convenient yet unjustified assumption that the reward function can be expressed
as a linear combination of features. One proven way to abandon this assumption
is by representing the reward function as a Gaussian process (GP)
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11,DBLP:journals/corr/abs-1208-2112}.
The original approach used maximum likelihood estimation
\cite{DBLP:conf/nips/LevinePK11}, while the goal of this project is to use
variational inference (VI) instead, which learns approximate posterior
probability distributions instead of point estimates. This approach can prove
useful in three major ways:
\begin{enumerate}
\item Modelling full posterior distributions for various parameters can result
  in more precise reward predictions, as the model simply holds more
  information.
\item Having variance estimates for rewards can direct our choice in what data
  should be collected next.
\item An approximate Bayesian treatment of many parameters in the model guards
  against overfitting \cite{DBLP:conf/uai/JinDAS17}.
\end{enumerate}

@@

\begin{definition}[MDP]
  A \emph{Markov decision process} is a set $\mathcal{M} = \{ \mathcal{S},
  \mathcal{A}, \mathcal{T}, \gamma, r \}$, where $\mathcal{S}$ and
  $\mathcal{A}$ are sets of states and actions, respectively; $\mathcal{T} :
  \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to [0, 1]$ is a function
  defined so that $\mathcal{T}(s, a, s')$ is the probability of moving to state $s'$
  after taking action $a$ in state $s$; $\gamma \in [0, 1)$ is the discount
  factor (with higher $\gamma$ values, it makes little difference whether a
  reward is received now or later, while with lower $\gamma$ values the future
  becomes gradually less and less important); and $r : \mathcal{S} \to
  \mathbb{R}$ is the reward function.
\end{definition}

In \emph{inverse reinforcement learning}, one is presented with an MDP without a
reward function $\mathcal{M} \setminus \{ r \}$ and a set of expert
demonstrations $\mathcal{D} = \{ \zeta_i \}_{i=1}^N$, where each demonstration
$\zeta_i = \{ (s_{i,0}, a_{i,0}), \dots, (s_{i,T}, a_{i,T}) \}$ is a multiset of
state-action pairs representing the actions taken by the expert during a
particular recorded session. Each state is also characterised by a number of
features. The goal of IRL is then to find $r$ such that the optimal policy under
$r$
\[ \pi^* = \argmax_\pi \mathbb{E}\left[ \left. \sum_{t=0}^\infty \gamma^t r(s_t)
    \right| \pi \right] \]
matches the actions in $\mathcal{D}$.

Following previous work on GP IRL
\cite{DBLP:conf/nips/LevinePK11,DBLP:conf/uai/JinDAS17}, we use a maximum
entropy IRL model \cite{ziebart2008maximum}, under which we have that
\[ P(a | s) \propto \exp(Q_{\mathbf{r}}(s, a)), \]
where
\[
  Q_{\mathbf{r}}(s, a) = r(s) + \gamma\sum_{s' \in \mathcal{S}}
  \mathcal{T}(s, a, s')\V(s'),
\]
and $\V : \mathcal{S} \to \mathbb{R}$ is an MDP \emph{value function}, which
can be obtained by repeatedly applying the 'soft' version of the Bellman backup
operator until convergence:
\cite{DBLP:conf/nips/LevinePK11,supplementary_material}
\begin{equation} \label{eq:update_rule}
  \V(s) \coloneqq \log \sum_{a \in \mathcal{A}} \exp\left( r(s) + \gamma\sum_{s'
      \in \mathcal{S}} \mathcal{T}(s, a, s')\V(s') \right).
\end{equation}
The likelihood of the data can then be expressed as
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}
\begin{equation} \label{pDr}
  \begin{split}
    p(\mathcal{D} | r) &= \prod_{i=1}^N \prod_{t=1}^T p(a_{i,t} | s_{i,t}) \\
    &= \exp\left( \sum_{i=1}^N \sum_{t=1}^T Q_{\mathbf{r}}(s_{i,t}, a_{i,t}) - \V(s_{i,t}) \right).
  \end{split}
\end{equation}
However, a reward function learned by maximising this likelihood is not
transferable to new situations
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}. One needs to model the
reward structure in a way that would allow reward predictions for previously
unseen states.

One way to model rewards without assumptions of linearity is with a
\emph{Gaussian process} (GP). A GP is a collection of random variables, any
finite combination of which has a joint Gaussian distribution
\cite{DBLP:books/lib/RasmussenW06}. We write $r \sim \mathcal{GP}(0,
k)$ to say that $r$ is a GP with mean $0$ and covariance function
$k$, which uses a vector of hyperparameters $\bm\lambda$.
Covariance functions take two state feature vectors as input and quantify how
similar the two states are, in a sense that we would expect them to have similar
rewards.

As training a GP with $n$ data points has a time complexity of
$\mathcal{O}(n^3)$ \cite{DBLP:books/lib/RasmussenW06}, numerous approximation
methods have been suggested, many of which select a subset of data called
\emph{inducing points} and focus most of the training effort on them
\cite{DBLP:journals/corr/abs-1807-01065}. Let $\mathbf{X_u}$ be the matrix of
features at inducing states, $\mathbf{u}$ the rewards at those states, and
$\mathbf{r}$ a vector with $r(\mathcal{S})$ as elements. Then the full joint
probability distribution can be factorised as
\[
  \pfull = p(\mathbf{u}) \times p(\mathbf{r} | \mathbf{u}) \times p(\mathcal{D} | r),
\]
where
\begin{align*}
  p(\mathbf{u}) &= \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu) \\
                &= \frac{1}{(2\pi)^{m/2}|\Kuu|^{1/2}}\exp \left( -\frac{1}{2} \mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} \right) \\
                &= \exp\left(-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} - \frac{1}{2}\log|\Kuu| - \frac{m}{2}\log 2\pi\right)
\end{align*}
is the GP prior \cite{DBLP:books/lib/RasmussenW06}, where $m \in \mathbb{N}$ is
the number of inducing points. The GP posterior is a
multivariate Gaussian \cite{DBLP:conf/nips/LevinePK11}
\begin{equation} \label{eq:r}
  p(\mathbf{r} | \bm\lambda, \mathbf{X_u}, \mathbf{u}) =
  \mathcal{N}(\mathbf{r}; \Kru^\intercal\Kuu^{-1}\mathbf{u}, \Krr - \Kru^\intercal\Kuu^{-1}\Kru),
\end{equation}
and $p(\mathcal{D} | r)$ is as in Equation \ref{pDr}. The matrices such as
$\Kru$ are called \emph{covariance matrices} and are defined as
$[\Kru]_{i,j} = k(\mathbf{x}_{\mathbf{r},i},
\mathbf{x}_{\mathbf{u},j})$, where $\mathbf{x}_{\mathbf{r},i}$ and
$\mathbf{x}_{\mathbf{u},j}$ denote feature vectors for the $i$th state in
$\mathcal{S}$ and the $j$th state in $\mathbf{X_u}$, respectively
\cite{DBLP:conf/uai/JinDAS17}.

Given this model, data $\mathcal{D}$, and inducing feature matrix
$\mathbf{X_u}$, our goal is then to find optimal values of hyperparameters
$\bm\lambda$, inducing rewards $\mathbf{u}$, and the rewards for all relevant
states $\mathbf{r}$. While the previous paper to consider this IRL model
computed maximum likelihood (ML) estimates for $\bm\lambda$ and $\mathbf{u}$,
and made an assumption that $\mathbf{r}$ in Equation \ref{eq:r} has zero variance
\cite{DBLP:conf/nips/LevinePK11}, we aim to avoid this assumption and use
VI to approximate the full posterior distribution $\posterior$.
\emph{Variational inference} is an approximation technique for probability
densities \cite{blei2017variational}. Let $\approximation$ be our approximating
family of probability distributions for $\posterior$ with its own hyperparameter
vector $\bm\nu$. Then the job of VI algorithms is to optimise $\bm\nu$ in order
to minimise the \emph{Kullback-Leibler} (KL) divergence between the original
probability distribution and our approximation.  KL divergence (asymmetrically)
measures how different the two distributions are, and in this case can be
defined as \cite{blei2017variational}
\begin{align*}
  \DKL{\approximation}{\posterior} &= \mathbb{E}[\log\approximation - \log\posterior ] \\
                                   &= \mathbb{E}[\log\approximation - \log\pfull] \\
                                   &+ \mathbb{E}[\log p(\mathcal{D}, \mathbf{X_u})].
\end{align*}
The last term is both hard to compute and constant w.r.t. $\approximation$
\cite{blei2017variational}, so we can remove it from our optimisation objective.
The negation of what remains is often called the \emph{evidence lower bound}
(ELBO) and is defined as\footnote{Throughout the proposal, all integrals should
  be interpreted as definite integrals over the entire sample space.}
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L} &= \mathbb{E}\left[ \log \frac{\pfull}{\approximation} \right] \\
    &= \iiint \log \frac{\pfull}{\approximation} \approximation\dx.
  \end{split}
\end{equation}

@@

\subsection{Linear Algebra and Numerical Analysis}

@@

% TODO: how do we name this?
\section{The WizWoz System}
% solution
% TODO: express all lambdas as exponentials (update the derivatives or sth)

@@

For any matrix $\mathbf{A}$, we will use either $A_{i,j}$ or
$[\mathbf{A}]_{i,j}$ to denote the element of $\mathbf{A}$ in row $i$ and column
$j$. Moreover, we use $\tr(\mathbf{A})$ to denote its \emph{trace} and
$\adj(\mathbf{A})$ for its \emph{adjugate} (or \emph{classical adjoint}).

@@

We primarily think of rewards as a vector $\mathbf{r} \in
\mathbb{R}^{|\mathcal{S}|}$, but sometimes we use a function notation $r(s)$ to
denote the reward of a particular state $s \in \mathcal{S}$.

@@

Derivatives such as $\frac{\partial \Kuu}{\partial \lambda_i}$ can be expressed
as
\[
  \frac{\partial \Kuu}{\partial \lambda_i} = \frac{1}{\lambda_i}\Kuu
\]

if $i = 0$, and

\[
  \begin{split}
    \left[ \frac{\partial \Kuu}{\partial \lambda_i} \right]_{j,k} =
    k(\mathbf{x}_{\mathbf{u},j}, \mathbf{x}_{\mathbf{u},k})
    \left( \vphantom{\frac{1}{2}} \right. &-\frac{1}{2}(x_{\mathbf{u},j,i} -
    x_{\mathbf{u},k,i})^2 \\
    &- \mathbbm{1}[j \ne k]\sigma^2 \left. \vphantom{\frac{1}{2}} \right)
  \end{split}
\]
otherwise.

% TODO: define PDF before this point
\begin{lemma}[Derivatives of PDFs] \label{lemma:derivatives}
  \item $\frac{\partial q(\mathbf{u})}{\partial \bm\mu} =
    \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} + \bm\Sigma^{-\intercal})(\mathbf{u}
    - \bm\mu)$.
      $\frac{\partial q(\mathbf{u})}{\partial \bm\Sigma} =
      \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal}
      - \bm\Sigma^{-\intercal})$.
      $\frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} =
      q(\mathbf{u})(\bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1} -
      |\bm\Sigma|^{-1}\adj(\bm\Sigma))\mathbf{B}$.
  \item For $i = 0, \dots, d$,
        \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} &=
        \frac{1}{2}q(\mathbf{r} \mid \mathbf{u}) (|\bm\Gamma|^{-1}
          \tr(\mathbf{R} \adj(\bm\Gamma)) \\
          &- (\mathbf{r} -
          \mathbf{Su})^\intercal\bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}(\mathbf{r}
          - \mathbf{Su})),
\end{lemma}
%\begin{proof}
%  \leavevmode
%  \begin{enumerate}
%  \item
%    \[
%      \begin{split}
%        \frac{\partial q(\mathbf{u})}{\partial m} &=
%        q(\mathbf{u})\dm\left[-\frac{Q}{2}\right]
%        \\
%        &= -\frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} +
%        \bm\Sigma^{-\intercal})(\mathbf{u} - \bm\mu)\dm[\mathbf{u} -
%        \bm\mu] \\
%        &= \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} +
%        \bm\Sigma^{-\intercal})(\mathbf{u} - \bm\mu).
%      \end{split}
%    \]
%  \item
%    \begin{align}
%      \frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} &= \dB\left[\frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\exp \left( -\frac{Q}{2} \right)\right] \nonumber \\
%                                                         &= \dB\left[\frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\right]\exp \left( -\frac{Q}{2} \right) \nonumber \\
%                                                         &+ \frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\dB\left[\exp\left( -\frac{Q}{2} \right)\right] \nonumber \\
%                                                         &= -\frac{1}{2} q(\mathbf{u}) \left( |\bm\Sigma|^{-3/2} \frac{\partial |\bm\Sigma|}{\partial \mathbf{B}} + \frac{\partial Q}{\partial \mathbf{B}} \right). \label{eq:2_before}
%    \end{align}
%    For derivatives with respect to $\bm\Sigma$, we can refer to Petersen and
%    Pedersen \cite{petersen2008matrix}:
%    \begin{equation} \label{eq:partial_derivatives1}
%      \frac{\partial |\bm\Sigma|}{\partial \bm\Sigma} = |\bm\Sigma|\bm\Sigma^{-\intercal}, \quad \frac{\partial Q}{\partial \mathbf{B}} = -\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal},
%    \end{equation}
%    while we can use an online tool by Laue et
%    al.\footnote{\url{http://www.matrixcalculus.org/}}
%    \cite{DBLP:conf/nips/LaueMG18} for the remaining ones:
%    \begin{equation} \label{eq:partial_derivatives2}
%      \begin{gathered}
%        \frac{\partial |\bm\Sigma|}{\partial \mathbf{B}} =
%        (\adj(\mathbf{T}) + \adj(\bm\Sigma))\mathbf{B}, \\
%        \dB[(\mathbf{u} - \bm\mu)^\intercal\bm\Sigma^{-1}(\mathbf{u} -
%        \bm\mu)] = -(\mathbf{TUT} +
%        \bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1})\mathbf{B}.
%      \end{gathered}
%    \end{equation}
%    Substituting results from Equations \ref{eq:partial_derivatives1} and
%    \ref{eq:partial_derivatives2} back into Equation \ref{eq:2_before} gives:
%    \begin{alignat*}{3}
%      \frac{\partial q(\mathbf{u})}{\partial \bm\Sigma} &=
%      \frac{1}{2}q(\mathbf{u})(&&\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal}
%      - \bm\Sigma^{-\intercal}), \\
%      \frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} &=
%      \frac{1}{2}q(\mathbf{u}) \{&&|\bm\Sigma|^{-3/2}(\adj(\mathbf{T}) +
%      \adj(\bm\Sigma)) \\
%      & &&+ \mathbf{TUT} + \bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1} \} \mathbf{B}.
%    \end{alignat*}
%  \item Using a result by Rasmussen and Williams % TODO: update the proof
%                                % according to the new result
%    \cite{DBLP:books/lib/RasmussenW06},
%    \[
%      \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} =
%      q(\mathbf{r} \mid \mathbf{u}) \dl
%      \left[-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} -

>>>>

\documentclass{mpaper}
\usepackage{bm}
\usepackage{bbm}
\usepackage{mathtools}
\usepackage{thmtools, thm-restate}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{tikz}

\usetikzlibrary{shapes,bayesnet,arrows.meta}
\tikzstyle{arrow} = [-{Latex[length=2mm,width=2mm]}]

@@

\newcommand{\approximationS}{q(\mathbf{u}_s, \mathbf{r}_s)}
\newcommand{\pfullS}{p(\mathcal{D}, \mathbf{u}_s, \mathbf{r}_s)}

@@

\section{Introduction}
% problem description/statement

Inverse reinforcement learning (IRL)---a problem proposed by Russell in
1998 \cite{DBLP:conf/colt/Russell98}---asks us to find a reward function for a
Markov decision process (MDP) that best explains a set of given demonstrations.
IRL is important because reward functions can be hard to define manually
\cite{DBLP:conf/icml/PieterN04,DBLP:journals/corr/abs-1806-06877}, and rewards
are not entirely specific to a given environment, allowing one to reuse the same
reward structure in previously unseen environments
\cite{DBLP:journals/corr/abs-1806-06877,DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}.
Moreover, IRL has seen a wide array of applications in autonomous vehicle
control \cite{DBLP:journals/ijsr/KimP16,DBLP:journals/ijrr/KretzschmarSSB16} and
learning to predict another agent's behaviour
\cite{DBLP:journals/ai/BogertD18,DBLP:conf/aaai/VogelRGR12,ziebart2008maximum,DBLP:conf/huc/ZiebartMDB08,DBLP:conf/iros/ZiebartRGMPBHDS09}.
Most approaches in the literature (see Section \ref{sec:background}) make a
convenient yet unjustified assumption that the reward function can be expressed
as a linear combination of features. One proven way to abandon this assumption
is by representing the reward function as a Gaussian process (GP)
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11,DBLP:journals/corr/abs-1208-2112}.
The original approach used maximum likelihood estimation
\cite{DBLP:conf/nips/LevinePK11}, while the goal of this project is to use
variational inference (VI) instead, which learns approximate posterior
probability distributions instead of point estimates. This approach can prove
useful in three major ways:
\begin{enumerate}
\item Modelling full posterior distributions for various parameters can result
  in more precise reward predictions, as the model simply holds more
  information.
\item Having variance estimates for rewards can direct our choice in what data
  should be collected next.
\item An approximate Bayesian treatment of many parameters in the model guards
  against overfitting \cite{DBLP:conf/uai/JinDAS17}.
\end{enumerate}

@@

\begin{definition}[MDP]
  A \emph{Markov decision process} is a set $\mathcal{M} = \{ \mathcal{S},
  \mathcal{A}, \mathcal{T}, \gamma, \mathbf{r} \}$, where $\mathcal{S}$ and
  $\mathcal{A}$ are sets of states and actions, respectively; $\mathcal{T} :
  \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to [0, 1]$ is a function
  defined so that $\mathcal{T}(s, a, s')$ is the probability of moving to state $s'$
  after taking action $a$ in state $s$; $\gamma \in [0, 1)$ is the discount
  factor; and $\mathbf{r} \in \mathbb{R}^{|\mathcal{S}|}$ is the reward
  vector\footnote{Depending on the situation, we will sometimes represent
    rewards as a function $r : \mathcal{S} \to \mathbb{R}$.}.
\end{definition}

\begin{definition}[IRL]
  Given an MDP without rewards $\mathcal{M} \setminus \{ \mathbf{r} \}$, an
  $|\mathcal{S}| \times d$ feature matrix $\mathbf{X}$ (where $d$ is the number
  of features), and a set of expert demonstrations $\mathcal{D} = \{\zeta_i
  \}_{i=1}^N$, where each demonstration $\zeta_i = \{ (s_{i,t}, a_{i,t})
  \}_{t=1}^T$ is a multiset of state-action pairs representing optimal actions
  executed by an expert, find the reward function that maximises the probability
  of observing the demonstrations, i.e.,
  \[
    \argmax_{\mathbf{r}} p(\mathcal{D} \mid \mathbf{r}).
  \]
\end{definition}

The optimal \emph{policy} $\pi : \mathcal{S} \to \mathcal{A}$ (i.e., a choice of
actions for each state that maximises reward over time) is usually constructed
by defining a \emph{value (utility) function} $\V : \mathcal{S} \to \mathbb{R}$
that measures how good a state is based on the reward $\mathbf{r}$ as well as
the structure of the MDP. One can then find $\V$ by applying the Bellman backup
operator until convergence to every $s \in \mathcal{S}$ (the technique is known
as \emph{value iteration}) \cite{DBLP:books/daglib/0023820}:
\[
  V_{\mathbf{r}}(s) \coloneqq r(s) + \gamma \max_{a \in \mathcal{A}} \sum_{s' \in
    \mathcal{S}} \mathcal{T}(s, a, s')V_{\mathbf{r}}(s').
\]

However, we follow previous work on GP IRL
\cite{DBLP:conf/nips/LevinePK11,DBLP:conf/uai/JinDAS17}, and use a
\emph{linearly solvable} (or \emph{maximum causal entropy}) MDP with stochastic
policy that define probability distributions over actions (instead of
suggesting a single action for each state) \cite{ziebart2008maximum}. This type
of MDP can be solved by applying the 'soft' version of the operator
\cite{DBLP:conf/nips/LevinePK11,supplementary_material}:
\begin{equation} \label{eq:update_rule}
  \V(s) \coloneqq \log \sum_{a \in \mathcal{A}} \exp\left( r(s) + \gamma\sum_{s'
      \in \mathcal{S}} \mathcal{T}(s, a, s')\V(s') \right).
\end{equation}
With this model, we can express the likelihood as
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}
\begin{equation} \label{pDr}
  \begin{split}
    p(\mathcal{D} \mid \mathbf{r}) &= \prod_{i=1}^N \prod_{t=1}^T p(a_{i,t} \mid s_{i,t}) \\
    &= \exp\left( \sum_{i=1}^N \sum_{t=1}^T Q_{\mathbf{r}}(s_{i,t}, a_{i,t}) - \V(s_{i,t}) \right),
  \end{split}
\end{equation}
where
\[
  Q_{\mathbf{r}}(s, a) = r(s) + \gamma\sum_{s' \in \mathcal{S}}
  \mathcal{T}(s, a, s')\V(s').
\]

However, a reward function learned by maximising this likelihood is not
transferable to new situations
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}. One needs to model the
reward structure in a way that would allow reward predictions for previously
unseen states.

One way to model rewards without assumptions of linearity is with a
\emph{Gaussian process} (GP). A GP is a collection of random variables, any
finite combination of which has a joint Gaussian distribution
\cite{DBLP:books/lib/RasmussenW06}. We write $r \sim \mathcal{GP}(0,
k)$ to say that $r$ is a GP with mean $0$ and covariance function
$k$. Covariance functions take two state feature vectors as input and quantify
how similar the two states are, in a sense that we would expect them to have
similar rewards.

As training a GP with $n$ data points has a time complexity of
$\mathcal{O}(n^3)$ \cite{DBLP:books/lib/RasmussenW06}, numerous approximation
methods have been suggested, many of which select a subset of data called
\emph{inducing points} and focus most of the training effort on them
\cite{DBLP:journals/corr/abs-1807-01065}. Let $\mathbf{X_u}$ be the matrix of
features at inducing states, $\mathbf{u}$ the rewards at those states. Then the
full joint probability distribution can be factorised as
\begin{equation} \label{eq:full}
  \pfull = p(\mathbf{u}) \times p(\mathbf{r} \mid \mathbf{u}) \times
  p(\mathcal{D} \mid \mathbf{r}),
\end{equation}
where
\begin{align*}
  p(\mathbf{u}) &= \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu) \\
                &= \frac{1}{(2\pi)^{m/2}|\Kuu|^{1/2}}\exp \left( -\frac{1}{2} \mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} \right) \\
                &= \exp\left(-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} - \frac{1}{2}\log|\Kuu| - \frac{m}{2}\log 2\pi\right)
\end{align*}
is the GP prior \cite{DBLP:books/lib/RasmussenW06}, and $m \in \mathbb{N}$ is
the number of inducing points. The GP posterior is then a multivariate Gaussian
\cite{DBLP:conf/nips/LevinePK11}
\begin{equation} \label{eq:r}
  p(\mathbf{r} \mid \mathbf{u}) =
  \mathcal{N}(\mathbf{r}; \Kru^\intercal\Kuu^{-1}\mathbf{u}, \Krr - \Kru^\intercal\Kuu^{-1}\Kru),
\end{equation}
and $p(\mathcal{D} \mid \mathbf{r})$ is as in Equation \ref{pDr}. The matrices such as
$\Kru$ are called \emph{covariance matrices} and are defined as
$[\Kru]_{i,j} = k(\mathbf{x}_{\mathbf{r},i},
\mathbf{x}_{\mathbf{u},j})$, where $\mathbf{x}_{\mathbf{r},i}$ and
$\mathbf{x}_{\mathbf{u},j}$ denote feature vectors for the $i$th state in
$\mathcal{S}$ and the $j$th state in $\mathbf{X_u}$, respectively
\cite{DBLP:conf/uai/JinDAS17}.

Given this model, data $\mathcal{D}$, and inducing feature matrix
$\mathbf{X_u}$, our goal is then to find optimal values of parameters
$\bm\lambda$, inducing rewards $\mathbf{u}$, and the rewards for all relevant
states $\mathbf{r}$. While the previous paper to consider this IRL model
computed maximum likelihood (ML) estimates for $\bm\lambda$ and $\mathbf{u}$,
and made an assumption that $\mathbf{r}$ in Equation \ref{eq:r} has zero variance
\cite{DBLP:conf/nips/LevinePK11}, we aim to avoid this assumption and use
VI to approximate the full posterior distribution $\posterior$.
\emph{Variational inference} is an approximation technique for probability
densities \cite{blei2017variational}. Let $\approximation$ be our approximating
family of probability distributions for $\posterior$. Then the job of VI is to
optimise the parameters of the approximating distribution in order to minimise
the \emph{Kullback-Leibler} (KL) divergence between the original probability
distribution and our approximation.  KL divergence (asymmetrically) measures how
different the two distributions are, and can be defined as
\cite{blei2017variational}
\begin{align*}
  \DKL{\approximation}{\posterior} &= \mathbb{E}[\log\approximation - \log\posterior ] \\
                                   &= \mathbb{E}[\log\approximation - \log\pfull] \\
                                   &+ \mathbb{E}[\log p(\mathcal{D})].
\end{align*}
The last term is both hard to compute and constant w.r.t. $\approximation$
\cite{blei2017variational}, so we can remove it from our optimisation objective.
The negation of what remains is often called the \emph{evidence lower bound}
(ELBO) and is defined as\footnote{Throughout the proposal, all integrals should
  be interpreted as definite integrals over the entire sample space.}
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L} &= \mathbb{E}\left[ \log \frac{\pfull}{\approximation} \right] \\
    &= \iiint \log \frac{\pfull}{\approximation} \approximation\dx.
  \end{split}
\end{equation}

@@

\subsection{Linear Algebra and Analysis}

@@

\section{Solution}
% TODO: express all lambdas as exponentials (update the derivatives or sth)

@@

We use $\tr(\mathbf{A})$ to denote its \emph{trace} and
$\adj(\mathbf{A})$ for its \emph{adjugate} (or \emph{classical adjoint}).

@@

@@

% TODO: define PDF before this point
\begin{restatable}[Derivatives of PDFs]{lemma}{derivatives} \label{lemma:derivatives}
  \leavevmode
  \begin{enumerate}
  \item $\frac{\partial q(\mathbf{u})}{\partial \bm\mu} =
    \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} + \bm\Sigma^{-\intercal})(\mathbf{u}
    - \bm\mu)$.
  \item
    \begin{enumerate}
    \item
      $\frac{\partial q(\mathbf{u})}{\partial \bm\Sigma} =
      \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal}
      - \bm\Sigma^{-\intercal})$.
    \item
      $\frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} =
      q(\mathbf{u})(\bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1} -
      |\bm\Sigma|^{-1}\adj(\bm\Sigma))\mathbf{B}$.
    \end{enumerate}
  \item For $i = 0, \dots, d$,
    \begin{enumerate}
    \item
      \begin{align*}
        \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} &= \frac{1}{2}q(\mathbf{r} \mid \mathbf{u}) (|\bm\Gamma|^{-1} \tr(\mathbf{R} \adj(\bm\Gamma)) \\
                                                                          &- (\mathbf{r} - \mathbf{Su})^\intercal\bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}(\mathbf{r} - \mathbf{Su})).
      \end{align*}
    \item For any covariance matrix $\mathbf{K}$,
      \[
        \frac{\partial \mathbf{K}}{\partial \lambda_i} =
        \begin{cases}
          \frac{1}{\lambda_i}\mathbf{K} & \text{if } i = 0, \\
          \mathbf{L} & \text{otherwise,}
        \end{cases}
      \]
      where
      \[
        L_{j,k} = k(\mathbf{x}_j, \mathbf{x}_k) \left( -\frac{1}{2}(x_{j,i} -
          x_{k,i})^2 - \mathbbm{1}[j \ne k]\sigma^2 \right).
      \]
    \end{enumerate}
  \end{enumerate}
\end{restatable}

\subsection{Details of the Model} \label{sec:details}

In order to properly investigate the difference between variational inference
and ML estimation, we keep other parts of the model the same.
Namely, we set the covariance function to a version of the automatic relevance
detection kernel \cite{DBLP:conf/nips/LevinePK11,neal2012bayesian}
\begin{align*}
  k(\mathbf{x}_i, \mathbf{x}_j) = \lambda_0\exp\left( \vphantom{\frac{1}{2}} \right. &-\frac{1}{2}(\mathbf{x}_i - \mathbf{x}_j)^\intercal\bm\Lambda(\mathbf{x}_i - \mathbf{x}_j) \\
                                                     &- \mathbbm{1}[i \ne j]\sigma^2\tr(\bm\Lambda) \left. \vphantom{\frac{1}{2}} \right),
\end{align*}
where $\lambda_0$ is the overall `scale' factor for how similar or distant the
states are, $\bm\Lambda = \diag(\lambda_1, \dots, \lambda_d)$ is a diagonal
matrix that determines the relevance of each feature (where $d$ denotes the
number of features), $\mathbbm{1}$ is defined as
\[ \mathbbm{1}[b] = \begin{cases}
    1 & \mbox{if $b$ is true} \\
    0 & \mbox{otherwise},
  \end{cases} \]
and $\sigma^2$ is set to $10^{-2}/2$ (as the original paper noted that
the value makes little difference to the performance of the algorithm
\cite{DBLP:conf/nips/LevinePK11}). Our vector of parameters for the
covariance function is then $\bm\lambda = (\lambda_0, \dots,
\lambda_d)^\intercal$. % TODO: write about lambda having to be variational

TODO: introductory sentence for the following assumption:
\begin{equation} \label{eq:approximation}
  \approximation = q(\mathbf{u}) \times q(\mathbf{r} \mid \mathbf{u}),
\end{equation}
where
\begin{equation}
  q(\mathbf{u}) = \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S}),
\end{equation}
where $\bm\mu \in \mathbb{R}^{m}$ is the mean vector and the $m \times m$
positive semi-definite matrix $\bm\Sigma$ is the covariance matrix
\cite{DBLP:conf/nips/ChengB17,DBLP:journals/jmlr/HensmanDS17,DBLP:conf/aaai/HoangHL17}.

  \begin{figure}
    \centering
    \begin{tikzpicture}
      \node[latent] (u) {$\mathbf{u}$};
      \node[latent, right=of u] (r) {$\mathbf{r}$};
      \node[obs, right=of r] (D) {$\mathcal{D}$};
      \node[const, below=of r] (lambda) {$\bm\lambda$};
      \node[const, below=of u] (mu) {$\bm\mu$};
      \node[const, left=of mu] (Sigma) {$\bm\Sigma$};
      \edge {mu, Sigma} {u};
      \edge {lambda, u} {r};
      \edge {r} {D};
    \end{tikzpicture}
    \label{fig:graphical_model}
  \end{figure}

This gives us our variational parameters $\bm\nu = (\bm\mu, \mathbf{B},
\bm\lambda)^\intercal$. TODO: reference the figure

\subsection{Evidence Lower Bound} \label{sec:elbo}
